Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-20 submission/manuscript.Rmd
First pass at Abstract and Importance sections

---
title: "Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure

@@

@@

## Abstract (250 words)

Something clever

## Importance (150 words)

Something clever

>>>>

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

# Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure

@@

\vspace{10mm}

## Observation format

@@

## Abstract

The gut microbiome is thought to have a role in the development of colorectal cancer by protecting against and exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals that was previously subjected to 16S rRNA gene and metagenomic shotgun sequencing. We measured the fecal concentrations of acetate, propionate, isobutyrate, and butyrate within the cohort and found that there were no meaningful differences in their concentration and disease status. When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were only able to explain less than 10% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although it is likely that bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited value.

## Importance

Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors. Short chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties. We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors. We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status. Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations. These data indicate that there is not a meaningful link between the gut microbiome, SCFAs, and tumor burden.

### 2019-03-20 submission/manuscript.tex
First pass at Abstract and Importance sections

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
Microbiota, and Colorectal Tumors}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

\begin{document}
\maketitle

@@

Fecal short chain fatty acids are not predictive of colorectal cancer
status and cannot be predicted based on bacterial community structure

@@

@@

\hypertarget{abstract-250-words}{%
\subsection{Abstract (250 words)}\label{abstract-250-words}}

Something clever

\hypertarget{importance-150-words}{%
\subsection{Importance (150 words)}\label{importance-150-words}}

Something clever

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

\begin{document}

@@

\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
\section{Fecal short chain fatty acids are not predictive of colorectal
cancer status and cannot be predicted based on bacterial community
structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colorectal-cancer-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}

@@

\vspace{10mm}

\hypertarget{observation-format}{%
\subsection{Observation format}\label{observation-format}}

@@

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The gut microbiome is thought to have a role in the development of
colorectal cancer by protecting against and exacerbating inflammation.
Short chain fatty acids (SCFAs), including butyrate, have been shown to
have anti-inflammatory properties and are produced in large quantities
by colonic bacteria. We assessed whether there was an association
between fecal SCFA concentrations and the presence of colonic adenomas
or carcinomas in a cohort of individuals that was previously subjected
to 16S rRNA gene and metagenomic shotgun sequencing. We measured the
fecal concentrations of acetate, propionate, isobutyrate, and butyrate
within the cohort and found that there were no meaningful differences in
their concentration and disease status. When we incorporated these
concentrations into Random Forest classification models trained to
differentiate between people with normal colons and those with adenomas
or carcinomas, we found that they did not significantly improve the
ability of 16S rRNA gene or metagenomic gene sequence-based models to
classify individuals. Finally, we generated Random Forest regression
models trained to predict the concentration of each SCFA based on 16S
rRNA gene or metagenomic gene sequence data from the same samples. These
models performed poorly and were only able to explain less than 10\% of
the observed variation in the SCFA concentrations. These results support
the broader epidemiological data that questions the value of fiber
consumption for reducing the risks of colorectal cancer. Although it is
likely that bacterial metabolites may serve as biomarkers to detect
adenomas or carcinomas, fecal SCFA concentrations have limited value.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

Considering colorectal cancer is the third leading cancer-related cause
of death within the United States, there is a great need to detect
colorectal tumors early without invasive colonoscopy procedures and to
prevent the formation of tumors. Short chain fatty acids (SCFAs) are
often used as a surrogate for measuring gut health and for being
anti-carcinogenic because of their anti-inflammatory properties. We
evaluated the fecal SCFA concentration of a cohort of individuals with
varying colonic tumor burden who were previously analyzed to identify
microbiome-based biomarkers of tumors. We were unable to find an
association between SCFA concentration and tumor burden or use SCFAs to
improve our microbiome-based models of classifying people based on their
tumor status. Furthermore, we were unable to find an association between
the fecal community structure and SCFA concentrations. These data
indicate that there is not a meaningful link between the gut microbiome,
SCFAs, and tumor burden.
